Complementing theory on adolescent emotion regulation using  machine learning

### 2022-02-26 manuscript/manuscript.Rmd
update 2022-02-26 14:53:24

Another paradoxical finding was that substance use and other "problem behaviors",
like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
This suggests that drug use derives its high importance from higher-order interactions with other predictors.
It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
Future research should explore this issue further.

@@

## Reflections form

* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
* Some predictors show almost no marginal effect (e.g., father's age, drug use)
  + This suggests they might be important in **interactions**

>>>>

Another paradoxical finding was that substance use and other "problem behaviors",
like time spent with delinquent peers, had very low variable importance ratings in predicting emotion dysregulation.
Even the highest ranked problem behavior, drug use, had no distinguishable marginal effect.
This suggests that drug use derives its high importance from higher-order interactions with other predictors.
It might be the case, for example, that for some youth, drug use contributes to self-exploration and social relationships, whereas for others, it may be a form of escapism, self-medication, or even self-harm.
This is merely a hypothesis, and the present study cannot conclusively identify what interactions are at play.
Future research should explore this further.

@@

Correspondence between research interest vs empirical importance.
one of the most commonly studied correlates of emotion regulation in adolescence is parenting [@vanlissaMappingPhenomenaRelevant2021]

One emergent theme in the developmental literature is attention for fathers' role in child development [ref Keizer].
Authors have lamented the lack of attention to fathers, and the equation of parenting with mothering.
With this in mind, it is worth noteworthy that the predictive importance of mothers' parenting did not in any way outrank that of fathers' parenting.
In fact, constructive conflict resolution with fathers was the highest-ranked construct related to conflict with parents;
paternal power was the highest-ranked parenting construct,
and maternal and paternal support and socio-economic status were both ranked approximately equally, despite both constructs being historically gendered.
This highlights the importance of considering both parents as equally relevant in research and in daily life.

Linear statistical models are the default in our field.
Therefore, an interesting result is that most predictors had non-linear effects:
Out of the three values contrasted in the marginal dependence plots,
only the one corresponding to the least desirable conditions diverged from the others.
<!-- For instanhigh neuroticism, high interparental conflict, low constructive conflict resolution with fathers. -->
Our findings send an important signal that many of the associations between predictors and developmental outcomes might be better modeled non-linearly.
In particular, they suggest a notion of "good-enough" preconditions for the development of mature emotion regulation.
Most adolescents seem to follow trajectories characterized by low difficulties in emotion regulation - except those with the most disadvantage.
Some prior research has considered this notion of a "good-enough" environment [@vanlissaCostEmpathyParentadolescent2017; @poonGoodEnoughParental2017];
the consistency of our findings suggests that it may be more ubiquitous than previously thought.

It is also noteworthy that some predictors had relatively high predictive importance,
but almost no marginal effect (e.g., father's age, drug use, openness to experience).
Such predictors likely derive their importance from interactions.
For instance, drug use might be a vehicle for social status or self-harm;
older fathers might be an asset if they are involved in children's lives but a liability if they prioritize professional success;
openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
As the present method is not well-suited to identify what interaction drives a variable's importance,
these results should be taken as indications for potentially fruitful areas of future research.
